A Machine Profile Summary

### 2022-06-20 MachineSummary/MachineSummary.Rmd
minor changes

In 2021, we undertook an investigation of tools in R for nonlinear least squares
calculations under the Google Summer of Code initiative. In the process, we recognized 
that we would be comparing calculations
performed on **different computing environments** at our disposal. Some of these were
based on different hardware, while others were different combinations of given hardware
with alternative operating software. In one case a machine failed and had to be replaced,
underlining the need to be able to compare results from the defunct system with its
replacements.

@@

There may be multiple terms used to describe some of the information we want, as
well as different interpretations of some of the units of measure. For example,
storage for disks or memory has decimal and binary schemes of units. Thus 
(see https://en.wikipedia.org/wiki/Gigabyte) there is a movement to use, for example, Gigabyte or GB to mean usits of 1000^3 bytes, while the Gigibyte or Gib is 1024^3.

@@

For tracking work, it is important to have 
information on when the calculations were performed. 
Most computations should have a time/date stamp that is unlikely to be confused with any
other, at least in combination with a machine name and/or other tags. Here is one
suggestion based on the R `Sys.time()` function. Moreover, as we were working on opposite
sides of the globe, it could be important to either specify time zone or to work with a
single "clock" such as UTC. R provides tools to allow this, as in the following example.

```{r tdstamp}
#get a timestamp
tsstr <- format(Sys.time(), "%Y/%m/%d|%H:%M") # tsstr == time stamp string in form YYYYmmddHHMM
cat("Date and Time stamp:", tsstr, "\n")
# and here is a way to get the UTC time
tutc <- as.POSIXlt(Sys.time(), tz = "UTC")
print(tutc)
utctstmp <- format(tutc, "%Y/%m/%d|%H:%M")
print(utctstmp)
```

@@

- a machine name. While most systems (and Linux in particular) offer to let the user
  provide a machine name, there are generally defaults that many people accept, and
  these are often uninformative and may be non-unique. We note that VirtualBox with
  a Windows 10 guest machine gave the name DESKTOP-HF4CKVA. Where this name was 
  generated we are not sure. Settings / System allows "rename this PC". Clearly 
  we may want to extend the name if particular software is involved, as follows.

@@

- cpu (i.e., processor)
- number of cores
- operating cycle speed (this is sometimes specific to the cpu specified )

@@

```{r envrontest}
envron <- function(){
D1 <- as.numeric(1) # use variable to avoid constants when double precision invoked
E5 <- 10 # arbitrary scaling for additive equality tests
B9 <- 1E+35 # big number, not necessarily biggest possible
E6 <- 1 # initial value for radix 
E9 <- 1 # initial value for machine precision
D0 <- E6 
repeat{
  E9 <- E9 / 2 # start of loop to decrease estimated machine precision
  D0 <- E6 + E9 # force storage of sum into a floating-point scalar
  if (D0 <= E6) break
} # repeat reduction while (1+E9) > 1
E9 <- E9 * 2 # restore smallest E9 which gives (1 + E9) > 1
repeat {
   E6 <- E6 + 1 # try different radix values
   D0 <- E6 + E9
   if (D0 <= E6) break
} # until a shift is observed
J1 <- 1 # initial count of radix digits in mantissa
E9 <- 1 # use radix for exact machine precision
repeat {  
  E9 <- E9 / E6 # loop while dividing by radix
  J1 <- J1 + 1 # increment counter
  D0 <- D1 + E9 # add tp 1
  if (D0 <= D1) break # test and repeat until equality
}
E9 <- E9 * E6 # recover last value of machine precision
J1 <- J1 - 1 # and adjust the number of digits
mpvals<-list(eps = E9, radix = E6, ndigits = J1)
}
mp <- envron()
E9 <- mp$eps
E6 <- mp$radix
J1 <- mp$ndigits
cat("E9 = ",E9,"  -- the machine precision,  E9=MIN (X  1+X>1)\n")
cat( "E9*1E+16=", E9 * 1E+16, "\n")
cat( "E6 = ", E6,"  -- the radix of arithmetic", "\n")
cat( "J1 = ", J1,"  -- the number of radix digits in mantissa of FP numbers", "\n")
cat( "E6^(-J1+1) * 1E+16=", 1E+16*(E6^(-J1+1)), "\n")
cat(".Machine$double.eps=", .Machine$double.eps, "\n")
cat("From R  .Machine:\n")
cat("double.eps =", .Machine$double.eps, "\n", 
    "double.base =", .Machine$double.base, "\n", 
    "double.digits =", .Machine$double.digits, "\n")
```

>>>>

In 2021, we undertook an investigation of tools in R for nonlinear least squares
calculations under the Google Summer of Code program. In the process, we recognized 
that we would be comparing calculations
performed on **different computing environments** at our disposal. Some of these were
based on different hardware, while others were different combinations of given hardware
with alternative operating software. In one case a machine failed and had to be replaced,
underlining the need to be able to compare results from the defunct system with its
replacements.

@@

There may be multiple terms used to describe some of the information we want, as
well as different interpretations of some of the units of measure. For example,
storage for disks or memory has decimal and binary schemes of units. Thus 
(see https://en.wikipedia.org/wiki/Gigabyte) there is a movement to use, for example, Gigabyte or GB to mean units of $1000^3$ bytes, while the Gibibyte or GiB is $1024^3$` bytes.

@@

For tracking work, it is important to have 
information on when the calculations were performed. 
Most computations should have a time/date stamp that is unlikely to be confused with any
other, at least in combination with a machine name and/or other tags. Here is one
suggestion based on the R `Sys.time()` function. Moreover, as we were working on opposite
sides of the globe, it deemed important to either specify time zone or to work with a
single "clock" such as UTC. R provides tools to allow this, as in the following example.

```{r tdstamp}
# get a time stamp
tsstr <- format(Sys.time(), "%Y/%m/%d|%H:%M") # tsstr == time stamp string in form YYYY/mm/dd|HH:MM
cat("Date and Time stamp:", tsstr, "\n")
# and here is a way to get the UTC time
tutc <- as.POSIXlt(Sys.time(), tz = "UTC")
print(tutc)
utctstmp <- format(tutc, "%Y/%m/%d|%H:%M")
print(utctstmp)
```

@@

- a machine name. While most systems (and Linux in particular) offer to let the user
  provide a machine name, there are generally defaults that many people accept, and
  these are often uninformative and may be non-unique. We note that VirtualBox with
  a Windows 10 guest machine gave the name DESKTOP-HF4CKVA. We are not sure where this name was generated. Settings / System allows "rename this PC". Clearly 
  we may want to extend the name if particular software is involved, as follows.

@@

- cpu (i.e., processor)
- number of cores
- operating cycle speed (this is sometimes specific to the cpu specified)

@@

```{r envrontest}
envron <- function(){
D1 <- as.numeric(1) # use variable to avoid constants when double precision invoked
E5 <- 10 # arbitrary scaling for additive equality tests
B9 <- 1E+35 # big number, not necessarily biggest possible
E6 <- 1 # initial value for radix 
E9 <- 1 # initial value for machine precision
D0 <- E6 
repeat{
  E9 <- E9 / 2 # start of loop to decrease estimated machine precision
  D0 <- E6 + E9 # force storage of sum into a floating-point scalar
  if (D0 <= E6) break
} # repeat reduction while (1+E9) > 1
E9 <- E9 * 2 # restore smallest E9 which gives (1 + E9) > 1
repeat {
   E6 <- E6 + 1 # try different radix values
   D0 <- E6 + E9
   if (D0 <= E6) break
} # until a shift is observed
J1 <- 1 # initial count of radix digits in mantissa
E9 <- 1 # use radix for exact machine precision
repeat {  
  E9 <- E9 / E6 # loop while dividing by radix
  J1 <- J1 + 1 # increment counter
  D0 <- D1 + E9 # add tp 1
  if (D0 <= D1) break # test and repeat until equality
}
E9 <- E9 * E6 # recover last value of machine precision
J1 <- J1 - 1 # and adjust the number of digits
mpvals<-list(eps = E9, radix = E6, ndigits = J1)
}
mp <- envron()
E9 <- mp$eps
E6 <- mp$radix
J1 <- mp$ndigits
cat("E9 = ", E9, "  -- the machine precision,  E9=MIN (X  1+X>1)\n")
cat( "E9*1E+16=", E9 * 1E+16, "\n")
cat( "E6 = ", E6, "  -- the radix of arithmetic", "\n")
cat( "J1 = ", J1, "  -- the number of radix digits in mantissa of FP numbers", "\n")
cat( "E6^(-J1+1) * 1E+16=", 1E+16*(E6^(-J1+1)), "\n")
cat(".Machine$double.eps=", .Machine$double.eps, "\n")
cat("From R  .Machine:\n")
cat("double.eps =", .Machine$double.eps, "\n", 
    "double.base =", .Machine$double.base, "\n", 
    "double.digits =", .Machine$double.digits, "\n")
```
